Behavioural Analysis of Independent Value-Based Learning in Non-cooperative Games

### 2025-06-07 paper/paper.tex
paper v7

\author{
\name{Yang Li\textsuperscript{a}\thanks{CONTACT Yang Li. Email: yang4.li@uwe.ac.uk}, Marco Perez Hernandez\textsuperscript{a} and Mehmet Emin Aydin\textsuperscript{a}}
\affil{\textsuperscript{a} University of the West of England, Coldharbour Lane, Frenchay Campus, Bristol, UK}
}

@@

\begin{abstract}
Multi-agent reinforcement learning has received a lot of updates in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Independent value-based algorithms such as independent Q-learning have been one of the most widely studied methods in the field due to their simplicity and scalability. Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we discuss the feasibility of these algorithms in non-cooperative settings. As a result, we explain the conditions a game must satisfy for the algorithms to work. We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis. Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
\end{abstract}

@@

\section{Introduction}
Recent multi-agent reinforcement learning (MARL) research has mostly focused on fully cooperative games \cite{yu2020benchmarking, papoudakis2020benchmarking, zhu2024survey} where there is no conflict between agents as they share a common reward. If we are allowed to share information among the agents, a fully cooperative game reduces to a single-agent game where the collective agent tries to find the optimal joint action, given the joint observation, although the problem is still challenging in practice because of the computational complexity of the joint action brings. In a traditional single-agent setting, the agent needs to learn which action gives the best reward. A Pareto-optimal pure strategy is guaranteed to exist in fully cooperative finite games, meaning that the agents can always deterministically choose the optimal action that has the highest payoff.

Non-cooperative environments, however, involve conflicts that make learning more complicated. Agents may belong to different parties that would like to have private control to protect their own interests. Rational self-interested agents will have to fight for what is best for them, as increasing one agent's reward might lead to the decline of other's gains. We can no longer intuitively define the solution for a game as the optimal strategy resulting in the highest reward. One of the possible solution concepts is the Nash Equilibrium (NE). NEs describe the stable points in games, which are associated with the convergence to optimal behaviour of agents.

@@

There are many attempts to conquer games with MARL. One of which is independent value-based learning (IVBL). IVBL is a family of algorithms that models the multi-agent environments as single-agent environments, and they compute for each agent the expected reward of each action under each state. IVBL is simple and highly scalable because it is decentralized. It has been adopted for both cooperative \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative} and non-cooperative tasks \cite{bjornsson2009cadiaplayer, jiang2018q, qu2020distributed, kopacz2023evaluating}. This paper analyses the performance of IVBL in non-cooperative finite games. The analysis considers characteristics of the action space and the NEs of the game, combining both formal analysis and experimental evaluation in selected two-player non-cooperative games. The results show NEs with uniformly random policies is a necessary condition for stability convergence of IVBL algorithms in non-cooperative environments. These condition highly restricts the wider applicability of IVBL.

\subsection{Main contributions}
\begin{itemize}
    \item Formal analysis identifying necessary conditions for stability convergence of IVBL which can drive the choice of algorithms in non-cooperative finite games.
    \item Introducing a non-cooperative game that has the characteristics which challenge the agents from a game theory perspective aligned with our formal analysis.
    \item Experimental evaluation of IVBL algorithms in selected two-player scenarios in the game proposed above.
\end{itemize}

\subsection{Contents}
The remainder of the paper is organized as follows:
In the next section, we provide the literature review on the key features of non-cooperative games and the most popular MARL algorithms that is related to our work. We then state the formal definitions of games and other concepts discussed in this paper in section 3. Section 4 analyses the stability of IVBL in non-cooperative games. To verify our analysis with experimental results, a new game called Food Chain is introduced in section 5. The game will have the key features discussed in our analysis that bring specific challenges to MARL agents. Then, we present the design as well as the results of our experiments in section 6. Finally, draw our conclusion in the last section.

\section{Tackling Non-Cooperative Games}
This literature review focuses on key features of non-cooperative games, as well as the MARL approaches to solve these games.

\subsection{Non-cooperative Games}
In single-agent or fully cooperative environments, there is one reward function for all agents. Thus, the agents only need to learn which (joint) action has the highest payoff. In non-cooperative games, there may be no optimal joint action that satisfy all agents. Many solution concepts to non-cooperative games are discussed in game theory \cite{albrecht2024multi}, such as Pareto optimum, correlated equilibrium, and Nash equilibrium. John Nash is famous for his study on non-cooperative games and his concept of Nash Equilibrium (NE) \cite{nash2024non}. NE describes a stable situation where no player can benefit from deviating from their current strategy independently. NEs offer a framework to characterise and analyse stability in games. Stability is associated to the optimal outcome in non-cooperative games and hence the objective agents aim for. Generally, there may only be a mixed NE, meaning that players must adopt randomized action-selection policy consisting of probability distributions over the action space. This is the mixed policy problem.

Many fully competitive games are studied in the MARL community such as Pommerman \cite{resnick2018pommerman}, Go \cite{silver2017mastering}, StarCraft II \cite{vinyals2019grandmaster}, and Dota 2 \cite{berner2019dota}. There are also some general non-cooperative environments such as hide-and-seek \cite{baker2019emergent}, Neural MMO \cite{suarez2021neural}. PettingZoo and Melting Pot \cite{terry2021pettingzoo, leibo2021scalable} are two integrated MARL environments including many cooperative and non-cooperative games. \cite{kopacz2023evaluating} studied the non-cooperative game predator-and-prey in the PettingZoo library, which is closely related to the game defined in our work. However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games. In cooperative games, the reward will simply increase as the agents get better at the game. This is no longer the case in non-cooperative games where increasing one agent's outcome often results in the drop of another agent's return. FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.

\subsection{MARL Algorithms}
Independent Learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment. The agents are essentially unaware of the existence of each other. IL is popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}. Independent value-based learning (IVBL) algorithms, such as the classic independent Q-Learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling. IVBL algorithms including IQL are often considered a baseline and a natural choice in a wide range of MARL studies because of their simplicity and scalability. For example, they were studied in various cooperative environments \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative}. The problem is that IVBL agents only give us the average reward of each action for each agent, but do not tell us what the action-selection policy look like. This is not a problem in single-agent or cooperative environments as we can simply pick the action with the highest return. Some studies addressed this issue in non-cooperative games by simply choosing the action with the highest empirical return \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating} and observed reasonably good outcome. \cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey. Both the algorithm and the game are closely related to our work. Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function (\ref{eq:softmax}) to calculate a policy based on the payoff. Softmax takes as input a vector and computes a probability distribution over the original vector elements, such that the higher the original value, the higher the probability.
\begin{equation}
    \textbf{softmax}([x_1, ..., x_n]) = [\frac{e^{x_1}}{\sum_{i}{e^{x_i}}}, ..., \frac{e^{x_n}}{\sum_{i}{e^{x_i}}}]
    \label{eq:softmax}
\end{equation}
The latter approach intuitively makes sense, as it emphasis the action with higher empirical return. However, these approaches have a huge problem that guarantees failure of learning in a wide range of games. We will take a closer look in the analysis section.

Unlike IL, joint action learning (JAL) algorithms allow the agent to gain access to the information processed by all other agents, such as their choices of actions, so agents can learn the expected return of joint actions. Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games. It computes local minimax solutions for each state based on the learned joint Q-values. Nash Q-learning \cite{hu2003nash} is a similar algorithm that computes the local NE instead. It converges to a NE under highly strict assumptions that reduce the applicability of the approach to a limited set of games. Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees. \cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games. Centralized learning requires full knowledge of the action spaces of all agents, which may not always be available if we are dealing with real-world problems. Another major disadvantage is that the number of joint actions is the number of possible combinations of the individual actions. This introduces combinatorics to the computational complexity of our algorithms, making them unscalable.

In contrast to the value-based approach, policy-based algorithms, such as the policy gradient methods, work around this issue by learning the policies directly, without putting an estimation on the individual action return. Policy gradient methods optimize a parametrized policy function using gradient decent to maximize the return. An advantage of policy gradient methods is that they naturally work on continuous action spaces. The actor-critic algorithms are a family of RL algorithm which trains a parametrized policy called the actor and a state-value function called the critic to estimate the policy gradient. Soft actor-critic (SAC) \cite{haarnoja2018soft} has become one of the most popular algorithms recently. Its multi-agent variant (MASAC) has been considered the state-of-the-art (SOTA) baseline \cite{yu2020benchmarking, zhang2020lyapunov}. There is also naturally an independent version (ISAC) for MARL. The proximal policy optimization algorithm (PPO) \cite{schulman2017proximal} was based on actor-critic as well and got extended to multi-agent environments by some researchers \cite{yu2020benchmarking, yu2022surprising} (MAPPO). Surprisingly, the IL version of PPO (IPPO) has proven to substantially outperform MAPPO and other algorithms in some cooperative tasks in StarCraft II \cite{de2020independent}. MADDPG \cite{lowe2017multi} is another policy gradient algorithm that is often considered SOTA. Because it only works for continuous action spaces, we do not discuss them further in the context of finite games.

@@

\subsection{Summary}
As shown in this review, the issue of deriving the policy from the Q-values learned by IVBL agents in non-cooperative games has not been fully addressed by the MARL research community, \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed} made progress in this direction, but they did not study the behaviour of the agents and identify the flaws of the algorithms from a game theory point of view.

\section{MARL Background}
In this section, we provide preliminaries including the formal definition of a game and NE.

\subsection{Stochastic Game}
A game is formally defined by a tuple:
$G = \langle N,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
\begin{itemize}
    \item $N = \{1, ..., n\}$: the set of agents, numbered from 1 to $n$.
    \item $S$: the game’s state space.
    \item $S_0 \subset S$: the set of possible initial states.
    \item $S_t \subset S$: the set of terminal states.
    \item $\{A_1,\ldots,A_n\}$: one action space for each agent. Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent. $R_i:S \times A^n \times S \rightarrow \mathbb{R}$.
    \item $T$: the state transition function. $T:S \times A^n \times S \rightarrow [0,1]$. It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
    \begin{equation}
        \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
        \label{eq:state_transition}
    \end{equation}
\end{itemize}
Starting from a randomly chosen $s_0 \in S_0$, at each time step $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$. The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent. If $s_t \in S_t$, the game ends. Otherwise, the game continues in the next time step.

The way a player plays the game defines a policy (strategy) as a function that computes the probability of selecting each action given the current state, $\pi:S \times A^n \rightarrow [0,1]$. A strategy is said to be pure if it is deterministic ($\pi:S \times A^n \rightarrow \{0,1\}$). Otherwise, we call it a mixed strategy. A set of strategies, one for each player, is called a profile $\pi^n$. Given the profile, we can compute the expected reward for each player in the game. The goal is to find the optimal strategy for an agent, or even the optimal profile, such that expected cumulative rewards are maximized. Cooperative games are special cases where agents share the same reward. The optimal profile is the one that maximizes the common reward. Because they share the same reward function, we can view the problem as a joint player selecting a joint action and receive a reward, which makes cooperative games essentially single-agent games from a game theory perspective. In general non-cooperative games, however, it may not be possible to make everyone happy as their are conflicting interests. A more profound definition of optimum need to be discussed in the non-cooperative games section.

Formally, let $R_i^*(\pi^n)$ be the function to compute the expected reward of the i'th player given the profile. Let $\pi_i$ be a policy of the i'th player and $\Pi_i$ be the policy space of the i'th player. Let $\pi^{-i}$ denote $\pi^n - \{\pi_i\}$, the set of strategies of all other players. A profile $\pi^n$ is a NE if
\begin{equation}
    \forall i \in N, \forall \pi_i \in \Pi_i: R_i^*(\{\pi_i\} \cup \pi^{-i}) \leqslant R_i^*(\pi^n)
    \label{eq:nash}
\end{equation}
A NE is called strict if $\leqslant$ is replaced by $<$. The idea is that NEs are the stable points in the game, because if an agent can benefit from switching strategy on its own, they would do so.

A game is finite if $|N|, |S|$ and all $|A_i|$ are finite, and the game always ends in a finite number of turns. John Nash proved that every finite game has at least one NE. To solve a game in the context of this paper means to find at least one NE. Games that are not finite can be ill-conditioned and have no NE. For example, a game where the player who names the greatest number wins has no best strategy. For the rest of the paper, we only refer to finite games. A normal form game has only two states, the starting state $S_0$ and the end state $S_1$. The game ends immediately after all agents pick an action. Thus, they can be represented using the reward matrix. In the book on game theory originally published in 1944 \cite{von2007theory}, Von Neumann and Morgenstern showed us that finite games can be translated into normal form games. If we encapsulate the following states as a subgame, we end up with a reward matrix for the initial state with the rewards being the expected rewards of the subgames. We focus on normal form games because they are easier to work with for the formal analysis. Since they only have one state $S_0$ for agents to take actions, we simply write $\pi(a)$ instead of $\pi(S_0, a)$ to denote the probability of taking action $a$.

\section{Stability of IVBL in Non-Cooperative Games}
In this part of the paper, we show the feasibility of IVBL algorithms in non-cooperative games with formal analysis. We demonstrate the necessary conditions the game must satisfy in order to apply the IVBL algorithms.

\begin{table}
    \begin{minipage}{0.5\textwidth}
    \tbl{Rock-Paper-Scissors Reward Matrix}{
    \begin{tabular}{l|ccc}
    & r & p & s\\
    \hline
    r & 0,0 & -1,1 & 1,-1 \\
    p & 1,-1 & 0,0 & -1,1 \\
    s & -1,1 & 1,-1 & 0,0 \\
    \end{tabular}
    }
    \label{table:rps_reward_matrix}
    \end{minipage}
    \begin{minipage}{0.5\textwidth}
    \tbl{Biased Rock-Paper-Scissors Reward Matrix}{
        \begin{tabular}{l|ccc}
        & r & p & s\\
        \hline
        r & 50,50 & 25,75 & 100,0 \\
        p & 75,25 & 50,50 & 45,55 \\
        s & 0,100 & 55,45 & 50,50 \\
        \end{tabular}
    }
    \label{table:biased_rps_reward_matrix}
    \end{minipage}
\end{table}

To demonstrate the problem with competitive games, let's analyse some simple normal form games. Consider the simple one-shot matrix game rock-paper-scissors (RPS) in Table~\ref{table:rps_reward_matrix}. Each entry represents the reward for the row player and the column player, respectively. The only NE in the game is that two players play each of the three actions with 1/3 chance. If any player deviates from the NE, their expected reward does not increase, and the other player can exploit them by playing the action that counters their most frequent action. Table~\ref{table:biased_rps_reward_matrix} defines a more complicated game called biased rock-paper-scissors. The only NE in the game is that both players adopt the following policy.
\begin{equation}
    \pi=(P(r)=0.0625,P(p)=0.6250,P(s)=0.3125)
    \label{eq:biased_rps_policy}
\end{equation}

>>>>

\author{
\name{Yang Li\textsuperscript{a}\thanks{CONTACT Yang Li. Email: yang4.li@uwe.ac.uk}, Marco Perez Hernandez\textsuperscript{a} and Mehmet Emin Aydin\textsuperscript{a}}
\affil{\textsuperscript{a} University of the West of England, Frenchay Campus, Coldharbour Lane, Bristol, UK}
}

@@

\begin{abstract}
Multi-agent reinforcement learning has received a lot of updates in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Independent value-based algorithms such as independent Q-learning have been one of the most widely studied methods in the field due to their simplicity and scalability. Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we discuss the feasibility of these algorithms in non-cooperative settings. As a result, we explain the conditions that a game must satisfy for the algorithms to work. We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our analysis. Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
\end{abstract}

@@

\section{Introduction}
Recent multi-agent reinforcement learning (MARL) research has mostly focused on fully cooperative games \cite{yu2020benchmarking, papoudakis2020benchmarking, zhu2024survey} where there is no conflict between agents as they share a common reward. When agents are allowed to share information, we can apply single-agent reinforcement learning (RL) algorithms to a fully cooperative game where the collective agent tries to find the optimal joint action, given the joint observation, although the problem is still challenging in practice because of what the computational complexity of the joint action brings. In a traditional single-agent setting, the agent needs to learn which action gives the best reward. A Pareto-optimal pure strategy is guaranteed to exist in fully cooperative finite games, meaning that the agents can always deterministically choose the optimal action that has the highest payoff.

Non-cooperative environments, however, involve conflicts that make learning more complicated. Agents may belong to different parties that would like to have private control to protect their own interests. Rational and self-interested agents will have to fight for what is best for them, as increasing one agent's reward might lead to the decline of other's gains. We can no longer intuitively define the solution for a game as the optimal strategy resulting in the highest reward. One of the possible solution concepts is the Nash equilibrium (NE). NEs describe the stable points in games, which are associated with the convergence to optimal behaviour of agents.

@@

Some studies have explored the application of MARL to solve games. One of which is independent value-based learning (IVBL). IVBL is a family of algorithms that models the multi-agent environments as single-agent environments, and they compute for each agent the expected reward of each action under each state. IVBL is simple and highly scalable because it is decentralized. It has been adopted for both cooperative \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative} and non-cooperative tasks \cite{bjornsson2009cadiaplayer, jiang2018q, qu2020distributed, kopacz2023evaluating}. This paper analyses the performance of IVBL in non-cooperative finite games. The analysis considers characteristics of the action space and the NEs of the game, combining both formal analysis and experimental evaluation in selected two-player non-cooperative games. The results show NEs with uniformly random policies is a necessary condition for stability convergence of IVBL algorithms in non-cooperative environments. These condition highly restricts the wider applicability of IVBL.

%\subsection{Main contributions}
The main contributions of this work include the following points:
\begin{itemize}
    \item Behavioural analysis identifying necessary conditions for a stable convergence of IVBL which can drive the choice of algorithms in non-cooperative finite games.
    \item Introducing a non-cooperative game that has the characteristics which challenge the agents from a game theory perspective aligned with our analysis.
    \item Experimental evaluation of IVBL algorithms in selected two-player scenarios in the game proposed above.
\end{itemize}

%\subsection{Contents}
The remainder of the paper is organized as follows:
The next section provides the literature review on the key features of non-cooperative games and the most popular MARL algorithms that is related to our work. We then state the formal definitions of games and other concepts discussed in this paper in Section 3. Section 4 analyses the stability of IVBL in non-cooperative games. To verify our analysis with experimental results, a new game called Food Chain is introduced in Section 5. The game will have the key features discussed in our analysis that bring specific challenges to MARL agents. Then, we present the design as well as the results of our experiments and relevant discussions in Section 6. Finally, the drawn conclusions pulled in the last section.

\section{Background and Literature Review}
This literature review focuses on key features of non-cooperative games, as well as the MARL approaches to solve these games.

\subsection{Non-Cooperative Games}
In single-agent or fully cooperative environments, there is one reward function for all agents. Thus, the agents only need to learn which (joint) action has the highest payoff. In non-cooperative games, there may be no optimal joint action that satisfy all agents. Several problem solving concepts/approaches for non-cooperative games are discussed in game theory \cite{albrecht2024multi}. These include Pareto optimum, Nash equilibrium, and correlated equilibrium. Nash Equilibrium (NE), proposed by John Nash, is a renown approach for non-cooperative games \cite{nash2024non}. NE offers a framework to characterise and analyse stability in games and describes a stability condition with which no player can benefit from deviating from their current strategy independently. The stability is associated to the optimal outcome in non-cooperative games and hence the objective agents aim for. Generally, there may only be a mixed NE, meaning that players must adopt randomized action-selection policy consisting of probability distributions over the action space. This is the mixed policy problem.

Many fully competitive games are studied in the MARL community such as Pommerman \cite{resnick2018pommerman}, Go \cite{silver2017mastering}, StarCraft II \cite{vinyals2019grandmaster}, and Dota 2 \cite{berner2019dota}. There are also some general non-cooperative environments such as hide-and-seek \cite{baker2019emergent}, Neural MMO \cite{suarez2021neural}. PettingZoo and Melting Pot \cite{terry2021pettingzoo, leibo2021scalable} are two integrated MARL environments including many cooperative and non-cooperative games. Kopazs et al. \cite{kopacz2023evaluating} studied the non-cooperative game predator-and-prey in the PettingZoo library, which is closely related to the game defined in this work. However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games. In cooperative games, the reward simply increases as the agents do better at the game. However, this is not the case in non-cooperative games where increasing one agent's outcome often results in the drop of another agent's return. FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.

\subsection{MARL Algorithms}
Independent learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment. The agents are essentially unaware of the existence of one another. IL algorithms are popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}. Among these, independent value-based learning (IVBL) algorithms, such as the classic independent Q-learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling. IVBL algorithms including IQL are often considered as baseline and natural choice in a wide range of MARL studies because of their simplicity and scalability. For example, they were studied in various cooperative environments \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative}. The problem is that IVBL agents only help produce the average reward of each action for each agent, but do not pay attention on what the action-selection policy. This is not a problem in single-agent or cooperative environments as we can simply pick the action with the highest return. Some studies addressed this issue in non-cooperative games by simply choosing the action with the highest empirical return \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating} and observed reasonably good outcome. Kopacz et al \cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey. Both the algorithm and the game are closely related to our work. Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function given in Equation~(\ref{eq:softmax}) to calculate a policy based on the payoff. Softmax takes a vector as input and computes a probability distribution over the original vector entries, such that the higher the original value, the higher the probability.
\begin{equation}
    \textbf{softmax}([x_1, ..., x_n]) = [\frac{e^{x_1}}{\sum_{i}{e^{x_i}}}, ..., \frac{e^{x_n}}{\sum_{i}{e^{x_i}}}]
    \label{eq:softmax}
\end{equation}
The latter approach emphasis the action with higher empirical return. However, these approaches have a huge problem that guarantees failure of learning in a wide range of games. We will take a closer look in the analysis section.

Unlike IL, joint action learning (JAL) algorithms allow the agent to gain access to the information processed by all other agents, such as their choices of actions, so agents can learn the expected return of joint actions. Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games. It computes local minimax solutions for each state based on the learned joint Q-values. Nash Q-learning \cite{hu2003nash} is a similar algorithm that computes the local NE instead. It converges to a NE under highly strict assumptions that reduce the applicability of the approach to a limited set of games. Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees. Zinkevich et al \cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games. Centralized learning requires full knowledge of the action spaces of all agents, which may not always be available if we are dealing with real-world problems. Another major disadvantage is that the number of joint actions is the number of possible combinations of the individual actions. This introduces combinatorics to the computational complexity of our algorithms, making them unscalable.

In contrast to the value-based approach, policy-based algorithms, such as the policy gradient methods, work around this issue by learning the policies directly, without putting an estimation on the individual action return. Policy gradient methods optimize a parametrized policy function using gradient decent to maximize the return. An advantage of policy gradient methods is that they naturally work on continuous action spaces. The actor-critic algorithms are a family of RL algorithm which trains a parametrized policy called the actor and a state-value function called the critic to estimate the policy gradient. Soft actor-critic (SAC) \cite{haarnoja2018soft} has become one of the most popular algorithms recently. Its multi-agent variant (MASAC) has been considered the state-of-the-art (SOTA) baseline \cite{yu2020benchmarking, zhang2020lyapunov}. There is also naturally an independent version (ISAC) for MARL. The proximal policy optimization algorithm (PPO) \cite{schulman2017proximal} was based on actor-critic as well and was extended to multi-agent environments by some researchers \cite{yu2020benchmarking, yu2022surprising} (MAPPO). Surprisingly, the IL version of PPO (IPPO) has proven to substantially outperform MAPPO and other algorithms in some cooperative tasks in StarCraft II \cite{de2020independent}. MADDPG \cite{lowe2017multi} is another policy gradient algorithm that is often considered SOTA. Because it only works for continuous action spaces, we do not discuss them further in the context of finite games.

@@

\subsection{Summary}
As shown in this review, the issue of deriving the policy from the Q-values learned by IVBL agents in non-cooperative games has not been fully addressed by the MARL community, \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed} made progress in this direction, but they did not analyse the behaviour of the agents and identify the flaws from a game theory point of view.

\section{MARL Preliminaries}
In this section, we provide preliminaries including the formal definition of a game and NE.

\subsection{Stochastic Game}
A game is formally defined by a tuple:

$G = \langle N,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
\begin{itemize}
    \item $N = \{1, ..., n\}$: the set of agents, numbered from 1 to $n$.
    \item $S$: the game’s state space.
    \item $S_0 \subset S$: the set of possible initial states.
    \item $S_t \subset S$: the set of terminal states.
    \item $\{A_1,\ldots,A_n\}$: one action space for each agent. Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent. $R_i:S \times A^n \times S \rightarrow \mathbb{R}$.
    \item $T$: the state transition function. $T:S \times A^n \times S \rightarrow [0,1]$. It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
    \begin{equation}
        \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
        \label{eq:state_transition}
    \end{equation}
\end{itemize}
Starting from a randomly chosen $s_0 \in S_0$, at each time step $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$. The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent. If $s_t \in S_t$, the game ends. Otherwise, the game continues in the next time step.

The way a player plays the game defines a policy (strategy) as a function that computes the probability of selecting each action given the current state, $\pi:S \times A^n \rightarrow [0,1]$. A strategy is said to be pure if it is deterministic ($\pi:S \times A^n \rightarrow \{0,1\}$). Otherwise, we call it a mixed strategy. A set of strategies, one for each player, is called a profile $\pi^n$. Given the profile, we can compute the expected reward for each player in the game. The goal is to find the optimal strategy for an agent, or even the optimal profile, such that expected cumulative rewards are maximized. Cooperative games are special cases where agents share the same reward. The optimal profile is the one that maximizes the common reward. Because they share the same reward function, the problem can be formulated as a joint player selecting a joint action and receives a reward, which makes cooperative games essentially single-agent games from a game theory perspective. In general non-cooperative games, however, it may not be possible to make everyone happy due to the conflicting interests. A more profound definition of optimum need to be discussed in the non-cooperative games section.

Formally, let $R_i^*(\pi^n)$ be the function to compute the expected reward of the $i^{th}$ player given the profile. Let $\pi_i$ be a policy of the $i^{th}$ player and $\Pi_i$ be the policy space of the $i^{th}$ player. Let $\pi^{-i}$ denote $\pi^n - \{\pi_i\}$, the set of strategies of all other players. A profile $\pi^n$ is a NE if
\begin{equation}
    \forall i \in N, \forall \pi_i \in \Pi_i: R_i^*(\{\pi_i\} \cup \pi^{-i}) \leqslant R_i^*(\pi^n)
    \label{eq:nash}
\end{equation}
The idea is that NEs are the stable points in the game, because if an agent can benefit from switching strategy on its own, they would do so.

A game is finite if $|N|, |S|$ and all $|A_i|$ are finite, and the game always ends in a finite number of turns. John Nash proved that every finite game has at least one NE. To solve a game in the context of this paper means to find at least one NE. Games that are not finite can be ill-conditioned and have no NE. For example, a game where the player who names the greatest number wins has no best strategy. For the rest of the paper, we only refer to finite games. A normal form game has only two states, the starting state $S_0$ and the end state $S_1$. The game ends immediately after all agents pick an action. Thus, they can be represented using the reward matrix. In the book on game theory originally published in 1944 \cite{von2007theory}, Von Neumann and Morgenstern showed that finite games can be translated into normal form games. If we encapsulate the following states as a subgame, we end up with a reward matrix for the initial state with the rewards being the expected rewards of the subgames. We focus on normal form games because they are easier to work with for the analysis. Since they only have one state $S_0$ for agents to take actions, we simply write $\pi(a)$ instead of $\pi(S_0, a)$ to denote the probability of taking action $a$.

\section{Stability of IVBL in Non-Cooperative Games}
In this part of the paper, we show the feasibility of IVBL algorithms in non-cooperative games with behavioural analysis. We demonstrate the necessary conditions the game must satisfy in order to apply the IVBL algorithms.

\begin{table}
    \begin{minipage}{0.5\textwidth}
    \tbl{Rock-Paper-Scissors Reward Matrix}{
    \begin{tabular}{l|ccc}
    & r & p & s\\
    \hline
    r & 0,0 & -1,1 & 1,-1 \\
    p & 1,-1 & 0,0 & -1,1 \\
    s & -1,1 & 1,-1 & 0,0 \\
    \end{tabular}
    }
    \label{table:rps_reward_matrix}
    \end{minipage}
    \begin{minipage}{0.5\textwidth}
    \tbl{Biased Rock-Paper-Scissors Reward Matrix}{
        \begin{tabular}{l|ccc}
        & r & p & s\\
        \hline
        r & 50,50 & 25,75 & 100,0 \\
        p & 75,25 & 50,50 & 45,55 \\
        s & 0,100 & 55,45 & 50,50 \\
        \end{tabular}
    }
    \label{table:biased_rps_reward_matrix}
    \end{minipage}
    \tabnote{The reward matrix is used to define the reward for each player for each choice of action. The row player chooses a row and gets the first reward. Vice versa for the column player.}
\end{table}

To demonstrate the problem with competitive games, we study two simple normal form games. Consider the simple one-shot matrix game rock-paper-scissors (RPS) in Table~\ref{table:rps_reward_matrix}. Each entry represents the reward for the row player and the column player, respectively. The only NE in the game is that two players play each of the three actions with 1/3 chance. If any player deviates from the NE, their expected reward does not increase, and the other player can exploit them by playing the action that counters their most frequent action. Table~\ref{table:biased_rps_reward_matrix} defines a more complicated game called biased rock-paper-scissors. The only NE in the game is that both players adopt the following policy.
\begin{equation}
    \pi=(P(r)=0.0625,P(p)=0.6250,P(s)=0.3125)
    \label{eq:biased_rps_policy}
\end{equation}

A well-known direct consequence of Equation~(\ref{eq:nash}) in game theory is that at any NE $\pi^n$, given the policy set of the other players $\pi^{i-1}$, player $i$ is indifferent from all actions that are chosen with a positive probability \cite{nash2024non}. This means that the actions are split into two sets, the worthy actions that are chosen with a positive probability and the worthless ones that will never be piked. For example, a variation of RPS could incorporate an always-losing action, where the player receives negative reward once executing it, regardless of the opponent’s action. Such an action should never be selected by the players. Additionally, the expected reward will be the same for every worthy action. In other words, the player can choose the worthy actions arbitrarily, given that other players commit to the NE. If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions indeed have the same payoff. The above fact can be formally stated as the following. Assume $\pi^n$ is a NE. Let $Q_i(a_j, \pi^{-i})$ be the expected reward of action $a_j \in A_i$ for agent $i$, given the rest of the profile $\pi^{-i}$. Let $A^+_i$ denote the set of worthy actions of agent $i$, namely $A^+_i = \{a | a \in A_i \wedge \pi_i(a) > 0 \}$. We have:
\begin{equation}
    \begin{aligned}
    \forall a_j, a_k \in A^+_i, Q_i(a_j, \pi^{-i}) = Q_i(a_k, \pi^{-i})
    \label{eq:nash_indifferent}
    \end{aligned}
\end{equation}

Suppose a IVBL algorithm can converge to a NE, then all worthy action's Q-values must converge to the same for an agent. It does not matter how the algorithm computes a probability distribution over the actions at the end, it cannot make any discriminations among those worthy actions based on equal statistics. Such an algorithm will output at best a uniform distribution over the worthy actions at a NE. This contradicts with the assumption that they converge to a NE unless the NE is full of uniformly random policies which is not the case in most of the games. We have shown that pure IVBL approaches does not work through proof by contradiction.
